---
output: hugodown::md_document
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Advent of Code 2020"
subtitle: ""
summary: "My attempts at [Advent of Code](https://adventofcode.com)."
authors: []
tags: []
categories: []
date: 2020-12-09
lastmod: 2020-12-09
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

[Advent of Code](https://adventofcode.com) is a series of small programming challenges, released daily throughout December in the run-up to Christmas. They are designed to be solved in any programming language. I will be using R. 

There will no doubt be a wide variety of ways of solving these problems. I'm going to go with the first thing I think of that gets an answer. In most cases, I image there will be more efficient solutions.

Each participant gets different input data, so my numerical solutions may be different from others.

1. <a href="#day1">Report repair</a>

<p><a id='day1'></a></p>

## Day 1: [Report Repair](https://adventofcode.com/2020/day/1)

The challenge is to find two numbers from a list that sum to 2020, then to report their product.

```{r message = FALSE}
library(dplyr)

expenses <- readLines("AoC_day1.txt") %>%
  as.numeric()

expand.grid(expenses, expenses) %>% 
  mutate(sum = Var1 + Var2) %>%
  filter(sum == 2020) %>%
  mutate(prod = Var1 * Var2) %>%
  slice(1) %>%
  pull(prod)
```

The follow-up challenge is the same but with three numbers. I went with essentially the same code but it's notably slower. There are a lot of repeated calculation here: each triplet appears six times in the table.

```{r}
expand.grid(expenses, expenses, expenses) %>% 
  mutate(sum = Var1 + Var2 + Var3) %>%
  filter(sum == 2020) %>%
  mutate(prod = Var1 * Var2 * Var3) %>%
  slice(1) %>%
  pull(prod)
```

