---
output: hugodown::md_document
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Testing custom fonts in plots"
subtitle: ""
summary: "Playing around with `ragg`"
authors: []
tags: []
categories: []
date: 2020-07-29
lastmod: 2020-07-29
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

## Testing using custom fonts with `ragg`

```{r}
library(ggplot2)
library(ragg)
```

### Using a font saved on my computer as .ttf

```{r}
ggplot(diamonds) + 
  geom_bar(aes(color, fill = color)) + 
  ggtitle("A fancy font") + 
  theme(text = element_text(family = "Neutraface Slab Display TT Bold", size = 20))
```
### Using a Google Font declared in `emk_font_set2`

```{r}
ggplot(diamonds) + 
  geom_bar(aes(color, fill = color)) + 
  ggtitle("A fancy font") + 
  theme(text = element_text(family = "Fira Code", size = 20))
```

### Using a variation of a Google Font saved in Font Book

```{r}
ggplot(diamonds) + 
  geom_bar(aes(color, fill = color)) + 
  ggtitle("A fancy font") + 
  theme(text = element_text(family = "FiraCode-Light", size = 10))
```

:::alert-note
- Using a Google Font NOT declared in `emk_font_set2` does not work.
- Also, using a .otf font in Font Book doesn't work.
- Also, using adding `face = "italic"` to `element_text` doesn't work if italic option not specified in font set.
:::


